---
title: "AD Mouse Processing"
author: "Alexander Bergendorf"
date: "2023-06-15"
output:
  pdf_document: default
  html_document: default
---



# Library Import
```{r}
library(Seurat) 
library(tidyverse) 
library(patchwork) 
library(dittoSeq) 
library(stringr)
library(gprofiler2)
library(harmony)
library(patchwork)
library(cowplot)
library(openxlsx)
library(MAST)

home_dir <- "C:/Users/aberg/OneDrive/Desktop/Brubaker Summer Research 2023"
```
# Normal Object Setup
## Data Import
```{r}
rownames <- rep("character", 1)
control <- rep("integer", 15013)
other <- rep("NULL", 46936)
colClasses <- c(rownames, control, other)
# Normal Cells
counts <- read.table(file = paste0(home_dir, "/AD_mouse_raw/GSE208683_Allcell_6m_UMI_matrix.txt"), sep = "\t", colClasses = colClasses)
metadata <- read.table(file = paste0(home_dir, "/AD_mouse_raw/GSE208683_Allcell_6m_meta_data.txt"), sep = "\t")

metadata <- metadata[1:15013,]
```


## Create Object
```{r}
colnames(counts) <- gsub("\\.", "-", colnames(counts))
wild_type <- CreateSeuratObject(counts = counts, meta.data = metadata)
table(wild_type@meta.data$SampleID, wild_type@meta.data$CellType)
rm(counts, metadata)
```


```{r}
saveRDS(wild_type, file = paste0(home_dir, "/AD_mouse_objects_ortholog/wt_ortholog.RData"))
#wild_type <- readRDS(file = paste0(home_dir, "/AD_mouse_objects_ortholog/wt_ortholog.RData"))
```


# AD Object Setup
## Data Import
```{r}
rownames <- rep("character", 1)
other1 <- rep("NULL", 31848)
ad <- rep("integer", 14662)
other2 <- rep("NULL",15439)
colClasses <- c(rownames, other1, ad, other2)
# Normal Cells
counts <- read.table(file = paste0(home_dir, "/AD_mouse_raw/GSE208683_Allcell_6m_UMI_matrix.txt"), sep = "\t", colClasses = colClasses)
metadata <- read.table(file = paste0(home_dir, "/AD_mouse_raw/GSE208683_Allcell_6m_meta_data.txt"), sep = "\t")
metadata <- metadata[31849:46510,]
```

## Create Object
```{r}
colnames(counts) <- gsub("\\.", "-", colnames(counts))
ad <- CreateSeuratObject(counts = counts, meta.data = metadata)
table(ad@meta.data$SampleID, ad@meta.data$CellType)
rm(counts, metadata)
```


```{r}
saveRDS(ad, file = paste0(home_dir, "/AD_mouse_objects_ortholog/ad_ortholog.RData"))
#ad <- readRDS(file = paste0(home_dir, "/AD_mouse_objects_ortholog/ad_ortholog.RData"))
```

# Merging and Processing Data
```{r}
ad <- readRDS(file = paste0(home_dir, "/ad_mouse_objects_ortholog/ad_ortholog.RData"))
wild_type <- readRDS(file = paste0(home_dir, "/ad_mouse_objects_ortholog/wt_ortholog.RData"))
ad <- SplitObject(ad, split.by = "SampleID")
wild_type <- SplitObject(wild_type, split.by = "SampleID")
combined <- c(ad, wild_type)
rm(ad, wild_type)
```

```{r}
table(combined_test@meta.data$SampleID)
```



```{r}
head(combined$AD_1@meta.data)
samples <- c("AD_1", "AD_2", "WT_1", "WT_2")
```

```{r}
mouse_ortholog <- readRDS(file = paste0(home_dir, "/AD_mouse_objects_ortholog/mouse_orthologs.RData"))
```


```{r}
for(i in samples){
  combined[[i]] <- combined[[i]][mouse_ortholog]
}
```

```{r}
for(i in samples){ 
  combined[[i]][["rna.size"]] <-   log10(Matrix::colSums(combined[[i]]@assays$RNA@counts)) 
} 
```




## QC and Visualization
```{r}
combined_test <- merge(x = combined[[1]], y = combined[2:length(combined)], merge.data=TRUE)
p1 <- VlnPlot(combined_test,"rna.size", group.by = "SampleID")
p1
p2 <- VlnPlot(combined_test,"percent.mt", group.by = "SampleID")
p2
```

```{r}
rna_size_min <- list() 
rna_size_max <- list() 

for(i in samples){ 
  rna_size_min[[i]] <- median(combined[[i]]@meta.data$rna.size) - (3*mad(combined[[i]]@meta.data$rna.size)) 
  
  rna_size_max[[i]] <- median(combined[[i]]@meta.data$rna.size) + (3*mad(combined[[i]]@meta.data$rna.size)) 
} 
```



```{r}
cairo_ps(filename= paste0(home_dir, "/Figures/Supplemental Figure 1/S12.eps"), width = 7, height = 7) # name the file as an EPS
p2 #variable of the figure that is defined/saved, can be used any name
dev.off()
```


```{r}
dittoPlot(combined_test, var = "nCount_RNA", group.by = "SampleID")

```

## Check Cell Cycle after clustering
```{r}
mmus_s <- gorth(cc.genes.updated.2019$s.genes, source_organism = "hsapiens", target_organism = "mmusculus")$ortholog_name
mmus_g2m <- gorth(cc.genes.updated.2019$g2m.genes, source_organism = "hsapiens", target_organism = "mmusculus")$ortholog_name
for(i in samples){ 
    combined[[i]] <- NormalizeData(combined[[i]]) 
    combined[[i]] <- CellCycleScoring(combined[[i]], s.features = mmus_s, g2m.features = mmus_g2m) 
}
```

## SCTransform
```{r}
#Applies SCTransform individually to each sample 

combined <- lapply(X = combined, FUN = SCTransform,  method ="glmGamPoi", return.only.var.genes = FALSE, vars.to.regress = c("S.Score", "G2M.Score")) 
```


```{r}
#saveRDS(combined, file = paste0(home_dir, "/AD_mouse_objects_ortholog/combined_ortholog.RData"))
combined <- readRDS(file = paste0(home_dir, "/AD_mouse_objects_ortholog/combined_ortholog.RData"))
```




```{r}
variable.features <- SelectIntegrationFeatures(object.list = combined, nfeatures = 3000)
```

## Batch
```{r}
combined <- merge(x = combined[[1]], y = combined[2:length(combined)], merge.data=TRUE) 

VariableFeatures(combined) <- variable.features 
```


```{r}
combined <- RunPCA(combined, verbose = FALSE) 
```

```{r}
stdev <- combined@reductions$pca@stdev
variances <- (stdev)^2 / sum(stdev^2)
variances
```



```{r}
#Checking differences in principal components 

options(repr.plot.height = 12, repr.plot.width = 20) 
p1 <- DimPlot(object = combined, reduction = "pca", pt.size = .1, group.by = "SampleID") 
#p2 <- VlnPlot(object = combined, features = "PC_1", group.by = "SampleID", pt.size = 0) 
#plot_grid(p1,p2)
cairo_ps(filename= paste0(home_dir, "/Figures/Supplemental Figure 1/1E_pre.eps"), width = 9, height = 7) # name the file as an EPS
p1 #variable of the figure that is defined/saved, can be used any name
dev.off()
```

We should batch correct.

```{r}
#Running Harmony  
options(repr.plot.height = 8, repr.plot.width = 16) 

combined <- RunHarmony(combined, assay.use = "SCT", group.by.vars = "orig.ident", plot_convergence = TRUE) 
ggsave(paste0(home_dir, "/Mouse_Viz/convergence_plot_ortholog.pdf"))
```

```{r}
p1 <- DimPlot(object = combined, reduction = "harmony", pt.size = .1, group.by = "SampleID") 

#p2 <- VlnPlot(object = combined, features = "harmony_1", group.by = "SampleID", pt.size = 0) 
cairo_ps(filename= paste0(home_dir, "/Figures/Supplemental Figure 1/1E_post.eps"), width = 9, height = 7) # name the file as an EPS
p1 #variable of the figure that is defined/saved, can be used any name
dev.off()
```

```{r}
stdev <- combined@reductions$harmony@stdev
variances <- (stdev)^2 / sum(stdev^2)
variances
```


```{r}
ElbowPlot(combined, reduction = "harmony", n = 50) 
```

Start with 30 PCS, check separation


## Reducing Dimensionality and Clustering
```{r}
#Performing Dimensional Reduction 

combined <- RunUMAP(combined, reduction = "harmony", dims = 1:30, verbose = FALSE) 

combined <- FindNeighbors(combined, reduction = "harmony", dims = 1:30, verbose = FALSE) 

combined <- FindClusters(combined, resolution = 1, verbose = FALSE) 
```


```{r}
dittoDimPlot(combined, "seurat_clusters", do.label = TRUE, legend.show = FALSE, labels.size = 3) 

dittoDimPlot(combined, "Genotype") 

dittoDimPlot(combined, "SampleID") 

dittoDimPlot(combined, "percent.mt") 

dittoDimPlot(combined, "CellType") 

dittoDimPlot(combined, "Phase") 
```

```{r}
#saveRDS(combined, file = paste0(home_dir, "/AD_mouse_objects_ortholog/mouse_adwt_final.RData"))
combined <- readRDS(file = paste0(home_dir, "/AD_mouse_objects_ortholog/mouse_adwt_final.RData"))
```




```{r}
marker_list <- c("Reln", "Satb2", "Fezf2", "Crym", "Bcl11b", "Sst", "Lhx6", "Adarb2", "Gad2", "Isl1", "Tcf7l2", "Eomes",
                 "Hes5", "Mki67", "Aldh1l1", "Olig2", "Otx2", "Trem2", "Igfbp7")

for(i in marker_list){
  print(dittoDimPlot(combined, var = i, assay = "RNA"))
  print(VlnPlot(combined, i, assay = "RNA"))
}
```

```{r}
dittoDimPlot(combined, "seurat_clusters", do.label = TRUE, legend.show = FALSE, labels.size = 3) 
```

```{r}
table(combined@meta.data$seurat_clusters, combined@meta.data$CellType)
```



## Finding Cluster Identifiers
```{r}
markers <- FindAllMarkers(combined, assay = "RNA", only.pos = TRUE, logfc.threshold = 0.5, test.use = "MAST", latent.vars = "SampleID") 

clust_counts <- table(combined@meta.data[,"seurat_clusters"]) 

clust_markers_list <- list() 

for(i in unique(markers$cluster)){ 

clustname <- paste0('cluster', i, '_N', clust_counts[i]) 

clust_markers_list[[clustname]] <- subset(markers, markers$cluster == i) 

rownames(clust_markers_list[[clustname]]) <- clust_markers_list[[clustname]]$gene 

} 

write.xlsx(clust_markers_list, file = paste0(home_dir, "/all_cell_markers_mouse_ortholog.xlsx"))
```

## Assign Idents
```{r}
Idents(combined) <- combined@meta.data$seurat_clusters 

new.cluster.ids <- c("Oligodendrocyte", "Neuron", "Astrocyte", "Neuron", "Neuron",
                     "Neuron", "Neuron", "Neuron", "Neuron", "Neuron",
                     "Neuron", "Microglia", "Neuron", "Oligodendrocyte Precursor", "Neuron",
                     "Neuron", "Neuron", "Endothelial", "Neuron", "Neuron",
                     "Neuron", "Neuron", "Astrocyte", "Neuron", "Astrocyte",
                     "Neuron", "Neuron", "Neuron", "Neuron", "Neuron",
                     "Neuron", "Endothelial", "Microglia", "Microglia", "Oligodendrocyte Precursor")

names(new.cluster.ids) <- levels(combined) 

combined <- RenameIdents(combined, new.cluster.ids) 

combined@meta.data$summary <- Idents(combined) 
```

```{r}
dittoDimPlot(combined, "summary")
dittoDimPlot(combined, "CellType")
dittoDimPlot(combined, "seurat_clusters", do.label = TRUE, legend.show = FALSE, labels.size = 3) 
```

```{r}
dittoDimPlot(combined, "Adcyap1")
VlnPlot(combined, "Adcyap1") + theme(legend.position = "none")
dittoDimPlot(combined, "Bdnf")
VlnPlot(combined, "Bdnf") + theme(legend.position = "none")
```


```{r}
Idents(combined) <- combined@meta.data$seurat_clusters 

new.cluster.ids <- c("Oligodendrocyte_0", "Neuron_1", "Astrocyte_2", "Neuron_3", "Neuron_4",
                     "Neuron_5", "Neuron_6", "Neuron_7", "Neuron_8", "Neuron_9",
                     "Neuron_10", "Microglia_11", "Neuron_12", "Oligodendrocyte Precursor_13", "Neuron_14",
                     "Neuron_15", "Neuron_16", "Endothelial_17", "Neuron_18", "Neuron_19",
                     "Neuron_20", "Neuron_21", "Astrocyte_22", "Neuron_23", "Astrocyte_24",
                     "Neuron_25", "Neuron_26", "Neuron_27", "Neuron_28", "Neuron_29",
                     "Neuron_30", "Endothelial_31", "Microglia_32", "Microglia_33", "Oligodendrocyte Precursor_34")

names(new.cluster.ids) <- levels(combined) 

combined <- RenameIdents(combined, new.cluster.ids) 

combined@meta.data$detail <- Idents(combined) 
```

```{r}
dittoDimPlot(combined, "summary")
dittoDimPlot(combined, "detail")
dittoDimPlot(combined, "CellType")
```
Source used for cell type assignment
https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6329831/
https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6671143/ 
https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6655561/


```{r}
table(combined@meta.data$summary, combined@meta.data$SampleID)
```



```{r}
saveRDS(combined, file = paste0(home_dir, "/AD_mouse_objects_ortholog/mouse_adwt_labeled.RData"))
#combined <- readRDS(file = paste0(home_dir, "/AD_mouse_objects_ortholog/mouse_adwt_labeled.RData"))
```

# Visualization
## 1D
```{r}
Idents(combined) <- combined@meta.data$summary
levels(combined) <- c("Astrocyte", "Endothelial", "Microglia", 
                      "Neuron", "Oligodendrocyte", "Oligodendrocyte Precursor")

p1 <- dittoDimPlot(combined, Idents(combined))

cairo_ps(filename= paste0(home_dir, "/Figures/Figure 1/1D.eps"), width = 9, height = 7) # name the file as an EPS
p1 #variable of the figure that is defined/saved, can be used any name
dev.off()
```

Remove 33 (Doublets)

## 1E
Markers
Endothelial: Igfbp7 (Cluster 17, 31)
Microglia: Trem2, Mrc1 (Cluster 11)
Oligodendrocyte Precursor: Matn4, Olig2, Pdgfr1 (Cluster 13, 34)
Astrocytes: Hes5, Aldh1l1 (Cluster 2, 22, 24)
Oligodendrocyte: Mbp, Mobp, Mog(Cluster 13, 34)
Neuron: Reln, Satb2, Fezf2, Crym, Bcl11b, Sst, Lnx6, Adarb2, Gad2 (Rest)

```{r}
markers <- c("Reln", "Satb2",
             "Mbp", "Mobp", "Mog", 
             "Matn4", "Pdgfra", 
             "Hes5", "Aldh1l1", 
             "Igfbp7",
             "Trem2", "Mrc1")
Idents(combined) <- combined@meta.data$summary
levels(combined) <- c("Neuron", "Oligodendrocyte", "Oligodendrocyte Precursor", "Astrocyte", "Endothelial", "Microglia")
combined <- ScaleData(combined, assay = "RNA")
p1 <- DoHeatmap(combined, markers, assay = "RNA", label = FALSE)

cairo_ps(filename= paste0(home_dir, "/Figures/Figure 1/1E.eps"), width = 9, height = 7) # name the file as an EPS
p1 #variable of the figure that is defined/saved, can be used any name
dev.off()
```

```{r}
Idents(combined) <- combined@meta.data$summary
levels(combined) <- c("Neuron", "Oligodendrocyte", "Oligodendrocyte Precursor", "Astrocyte", "Endothelial", "Microglia")
markers <- rev(markers)
p2 <- DotPlot(combined, features = markers, assay = "RNA") + 
  theme(axis.text.x = element_text(angle = 45, hjust=1))
cairo_ps(filename= paste0(home_dir, "/Figures/Figure 1/1E_dotplot.eps"), width = 9, height = 7) # name the file as an EPS
p2 #variable of the figure that is defined/saved, can be used any name
dev.off()
```



```{r}
dittoHeatmap(combined, markers, assay = "RNA", annot.by = c("summary", "seurat_clusters"), order.by = , cluster_rows = FALSE)
```

```{r}
levels(combined) <- c("Neuron", "Oligodendrocyte", "Oligodendrocyte Precursor", "Astrocyte", "Endothelial", "Microglia")
```

```{r}
cell_count_table <- table(combined@meta.data$summary, combined@meta.data$Genotype)
write.csv(cell_count_table, file = paste0(home_dir, "/Figures/Figure 1/mouse_counts.csv"))
```



```{r}
for(i in markers){
  print(i)
  dittoDimPlot(combined, i)
}
```

```{r}
sessionInfo()
```

